Bayesian calibration of a hybrid nuclear collision model\\ using $\boldsymbol p$-Pb and Pb-Pb data from the LHC

### 2018-10-12 paper/paper.tex
Save state before weekend

\newcommand{\trento}{T\raisebox{-0.5ex}{R}ENTo}
\newcommand{\sqrts}{\sqrt{s_\mathrm{NN}}}
\newcommand{\fmc}{\ensuremath{\text{fm}/c}}
\newcommand{\nch}{N_\text{ch}}
\newcommand{\ntrk}{N_\text{trk}^\text{offline}}
\newcommand{\vnk}[2]{v_#1\{#2\}}
\newcommand{\sigmaf}{\sigma_\text{fluct}}
\newcommand{\X}{\chi_\text{struct}}
\newcommand{\taufs}{\tau_\text{fs}}
\newcommand{\dmin}{d_\text{min}}
\newcommand{\Tsw}{T_\text{switch}}
\newcommand{\T}{\tilde{T}}
\newcommand{\x}{\mathbf x}
\newcommand{\y}{\mathbf y}
\newcommand{\z}{\mathbf z}
\newcommand{\paddedhline}{\noalign{\smallskip}\hline\noalign{\smallskip}}
\newcommand{\order}[1]{$\mathcal O(10^{#1})$}
\newcommand{\note}{\textcolor{theblue}{[?]}}

@@

Consider the collision of two protons, labeled $A$ and $B$, with three-dimensional densities $\rho_A$ and $\rho_B$ in their local rest frames.
The proton-proton overlap function
\begin{equation}
  \label{eq:tpp}
  T_{pp}(b) \equiv \int dx\, dy\, dz\, \rho_A(x, y, z)\, \rho_B(x - b, y, z),
\end{equation}
describes the eikonal overlap of the two proton wave packets at fixed impact parameter $b$, where coordinate $(x, y)$ lies in the transverse plane and $z$ is parallel to the beam axis.
Here we assume that each proton is comprised of smaller constituents---e.g.\ valence quarks, sea quarks, and small-x gluons---which may collide to produce secondary particles and contribute to the observed inelastic proton-proton cross section.

Within a picture of independent pairwise collisions between the constituents, a Glauber model model may be used to calculate the probability $P_\mathrm{coll}$ that the two protons collide inelastically at impact parameter $b$. In the limit when the number of constituents is large, it yields the particularly simple form
\begin{equation}
  \label{eq:pcoll}
  P_\mathrm{coll}(b) = 1 - \exp[-\sigma_\mathrm{eff}\, T_{pp}(b)], \\
\end{equation}
where $\sigma_\mathrm{eff}$ is an effective cross section for pairwise inelastic collisions between the constituents, and $\sigma_{pp}^\mathrm{inel}$ is the total inelastic proton-proton cross section
\begin{equation}
  \label{eq:sigma_nn}
  \sigma_{pp}^\mathrm{inel} = \int 2 \pi b\, db\, P_\mathrm{coll}(b).
\end{equation}
The proton densities $\rho_{A,B}$ in Eq.~\eqref{eq:tpp} are commonly modeled using a spherically symmetric distribution.
For instance, the original implementation of the \trento\ model uses Gaussian protons, largely because it yields a simple analytic solution to Eq.~\eqref{eq:pcoll}.
Needless to say, such approximations are admittedly crude and may have a significant effect on the dynamics of small collision systems where the proton size is comparable to the size of the produced QGP medium.

A number of previous studies have investigated the effects of deformed or ``lumpy'' protons.
One common implementation is a superposition of three valence quarks, typically described by Gaussian or exponential form factors \cite{Welsh:2016siu, Bozek:2017jog, Schenke:2014zha, Schlichting:2014ipa, Adare:2015bua, Broniowski:2016pvx}.
The corresponding proton density is then assumed to be that of predominantly small-x gluons, seeded by the distribution of color charge in each of the three valence quarks.

@@

In this work, we pursue a less restrictive and more parametric description of the proton where the number of substructure degrees of freedom is uncertain as depicted in Fig.~\ref{fig:substructure}.
We model each nucleon's three-dimensional density $\rho_{A,B}$ as a sum of $n_c$ independent constituents,
\begin{equation}
  \label{eq:rho}
  \rho_{A, B}(\x) = \frac{1}{n_c} \sum\limits_{i=1}^{n_c} \rho_c(\x - \x_i),
\end{equation}
where each constituent density $\rho_c$ is described by a Gaussian distribution,
\begin{equation}
  \rho_c(\x) = \frac{1}{(2 \pi v^2)^{3/2}} \exp\left(- \frac{|\x|^2}{2 v^2}\right).
\end{equation}
of width $v$.
The constituent positions $\x_i$ are sampled independently (ignoring correlations) from a Gaussian radial distribution,
\begin{equation}
  \label{eq:radial_dist}
  P(\x_i; \x_{A,B}) = \frac{1}{(2\pi r^2)^{3/2}} \exp\left(-\frac{|\x_i - \x_{A,B}|^2}{2 r^2}\right),
\end{equation}
where $\x_{A,B}$ is the position of each proton, and $r$ is a parameter which varies the sampling radius of the constituent positions about each proton position.
As a matter of convenience, we sample the proton positions \emph{before} determining the constituent positions.
This leads to small discrepancies between the designated proton positions $\x_{A,B}$, and the actual position of each proton's center-of-mass, owing to fluctuations in the constituent positions $\x_i$.
The sampling radius $r$ should thus be interpreted with care; it is \emph{not} the root-mean-square radius of the sampled nucleons.

@@

We assign the two protons $A$, $B$ a random impact parameter, and sample their inelastic collision probability using Eq.~\eqref{eq:pcoll}.
Assuming they collide, we assign each proton a fluctuated thickness function,
\begin{equation}
  \label{eq:fluctuated_thick}
  \T_{A, B}(x, y) \equiv \int dz\, \tilde{\rho}_{A,B}(x \pm b/2, y, z),
\end{equation}
which projects each fluctuated proton density $\tilde{\rho}_{A,B}$ onto the transverse plane $(x, y)$ centered at $z=0$.

Note, if any part of the proton collides, the entire proton contributes to inelastic particle production.
There is no concept of individually wounded constituents; the entire proton either collides (or misses) as a singular connected object.
This is an important distinction between the present model and a similar nucleon substructure implementation known as the participant or ``wounded'' quark model which allows for a subset of quarks (constituents) to participate inside a single nucleon \cite{ANISOVICH1978477, Broniowski:2016pvx}.
The proton, unlike the nucleus, cannot produce semi-stable spectator fragments in a high-energy collision.
Any spectator quarks produced by a wounded quark model would be colored objects that necessarily contribute to secondary particle production as they fragment and recombine to form color-neutral hadrons.
We correspondingly require that the nucleons in Eq.~\eqref{eq:rho} participate as singular objects, such that all spectator matter discarded by the simulation is appropriately color-neutral and inert.

The initial energy density $e$ deposited at midrapidity and proper time $\tau_0 \approx 0~\fmc$ is some function
\begin{equation}
  e(\x_\perp, \eta_s=0, \tau_0) = f(\T_A, \T_B)
\end{equation}
of the local density of participant matter $\T_A$, $\T_B$ in each nucleus.
A natural first guess for this mapping is the arithmetic mean
\begin{equation}
  f = \text{const} \times \frac{\T_A + \T_B}{2},
\end{equation}
which is simply a wounded nucleon model up to meaningless factor of two in the normalization.
The wounded nucleon model was in fact one of the first such mappings used as a proxy for initial particle production and entropy deposition in relativistic heavy-ion collisions \cite{Bialas:1976ed}.
It was quickly realized, however, that the wounded nucleon model predicts the wrong scaling for charged particle production as a function of collision centrality and hence the wrong scaling for initial entropy deposition as a function of participant thicknesses $\T_A$ and $\T_B$ \cite{Kharzeev:2000ph}.

A simple remedy is to replace the arithmetic mean of the wounded nucleon model with a more flexible parametrization
\begin{equation}
  \label{eq:gmean}
  \frac{dS}{d^2x_\perp d\eta}\,\bigg\vert_{\eta=0} \propto \bigg( \frac{\T_A^p + \T_B^p}{2} \bigg)^{1/p},
\end{equation}
based on a family of functions known as the generalized mean(s).
This parametrization introduces a dimensionless parameter $p$ which varies the scaling behavior of initial entropy deposition at midrapidity.
For certain discrete values of $p$, it reduces to well known functional forms such as the arithmetic, geometric, and harmonic means:
\begin{equation}
  \newlength{\extraspace}
  \setlength{\extraspace}{0.5ex}
  \frac{dS}{d\eta}\bigg\vert_{\eta=0} \propto
  \begin{cases}
    \max(T_A, T_B) & p \rightarrow +\infty, \\[\extraspace]
    (T_A + T_B)/2 & p = +1, \hfill \text{ (arithmetic)} \\[\extraspace]
    \sqrt{T_A T_B} & p = 0, \hfill \text{ (geometric)} \\[\extraspace]
    2\, T_A T_B/(T_A + T_B) & p = -1, \hfill \text{ (harmonic)} \\[\extraspace]
    \min(T_A, T_B) & p \rightarrow -\infty.
  \end{cases}
  \label{eq:trento_p}
\end{equation}
Conveniently, it has been shown that the generalized mean ansatz is able to mimic the scaling behavior of certain saturation based initial condition models \cite{Bernhard:2016tnd}, and hence it should serve as a reasonable parametric form for exploring QGP entropy deposition, assuming imperfect knowledge of saturation effects in nature.
Of course the model is \emph{not} a substitute for first principle theory calculations, and it may fail to reproduce nuanced features of \emph{ab initio} models such as the existence of short-range gluon field fluctuations \cite{Schenke:2012wb}.

@@

When applying the model to heavy-ions, we sample nucleon positions from a Woods-Saxon density distribution subject to a minimum distance criteria $|\x_i - \x_j| > d_\mathrm{min}$ between all pairs of nucleons $i$, $j$.
The minimum distance algorithm, first described in Ref.~\cite{Bernhard:2018hnz}, uses a simple trick to resample the nucleon positions without modifying the target Woods-Saxon radial distribution.
We first presample the radii of all nucleons in a given nucleus and sort them in ascending order.
We then sample the solid angles of each nucleon one-by-one, starting with the nucleon closest to the center of the nucleus and working our way outwards.
If a sampled nucleon position is too close to any of its previously placed neighbors, its solid angle is resampled until the minimum distance criteria is satisfied.
Similar methods could be used to model correlations between individual constituents inside each nucleon, although the numerical implementation would be somewhat tedious.

@@

The nuclear collision model constructed in Sec.~\ref{sec:model} includes a number of free parameters $\x$ which describe the initial state, pre-equilibrium dynamics, and hydrodynamic medium.
Given values for the parameters $\x$, the model may be used to predict a vector of simulated observables $\y_m$.
For example, $\y_m$ might be a vector consisting of charged particle yields in different centrality bins.
The physics model thus represents a vector-valued function $f(\x) = \y_m$ which maps the parameter values $\x$ to the simulated observables $\y_m$.

The goal of this work is to estimate the true model parameters $\x_\star$ provided some evidence that our model predictions $\y_m$ describe experimental measurements $\y_e$.
The problem involves three distinct components:
\begin{enumerate}[itemsep=0pt, leftmargin=2\parindent]
  \item $H_f$: the hypothesis that the nuclear collision model $f$ formulated in this work provides a realistic description of reality,
  \item $H_\x$: the hypothesis that the model parameters $\x$ are the true model parameters $\x_\star$ of $f$, and
  \item $E$: the evidence provided by the experimental data $\y_e$ and its corresponding uncertainties.
\end{enumerate}
As a practical matter, we always assume that hypothesis $H_f$ is correct, meaning there are no glaring flaws in our chosen theoretical framework.
This is a significant assumption; the application of hydrodynamic simulations to small collision systems is speculative, and our conclusions are conditional on the framework making sense.

Subject to this assumption, we can apply Bayes' theorem to evaluate the hypothesis $H_\x$ for the true model parameters,
\begin{equation}
  \label{eq:bayes}
  P(H_\x | E) \propto P(E | H_\x)\, P(H_\x).
\end{equation}
The left-side of this expression is the \emph{posterior}: the probability for hypothesis $H_\x$ given the experimental evidence $E$.
On the right-side there are two separate terms.
The first term $P(E | H_\x)$ is the \emph{likelihood} function: the probability of observing the experimental evidence $E$ given our model and the hypothesis $H_\x$ for the true model parameters $\x_\star$, and the second term $P(H_\x)$ is the \emph{prior}: an estimate of the probability of hypothesis $H_\x$ in the absence of evidence $E$.

We assume that the likelihood function in Eq.~\eqref{eq:bayes} is described by a multivariate normal distribution
\begin{equation}
  \label{eq:likelihood}
  P(E | H_\x) = \frac{1}{\sqrt{(2\pi)^m \det \Sigma}} \exp \left ( -\frac{1}{2}\Delta\y^\intercal \Sigma^{-1} \Delta\y \right ),
\end{equation}
where $\Delta\y = \y_m(\x) - \y_e$ is the discrepancy of the model and experiment, and $\Sigma = \Sigma_m(\x) + \Sigma_e$ is the \emph{total} covariance matrix, equal to the sum of a modeling component $\Sigma_m(\x)$ and an experimental component $\Sigma_e$ which account for all known sources of uncertainty in the simulated and measured observables.

@@

For the prior $P(H_\x)$, we specify ranges, i.e.\ minimum and maximum values, for each parameter which are listed in Table~\ref{tab:design}.
We assume the prior distribution is constant and nonzero within each specified range and zero otherwise.
The selected parameter ranges are intentionally wide to avoid clipping the calibrated posterior; for example, a previous analysis of the \trento\ model \cite{Bernhard:2016tnd} found $p \sim 0$, but we use a prior range $p \in [-1, 1]$ to account for differences in the present model, e.g.\ nucleon substructure, which could modify its posterior.
Several of the model parameters require special care and are reparametrized accordingly:

@@

The likelihood function \eqref{eq:likelihood} provides evidence for \mbox{(or against)} the model parameters $\x$ by comparing the model predictions $\y_m(\x)$ to experimental data $\y_e$.
We focus on simple experimental observables in the present study which are sensitive to the bulk properties of the produced medium.
We calculate for each set of model parameters the following observables at midrapidity:
\begin{itemize}[leftmargin=1\parindent, itemsep=0pt]
  \item Charged-particle multiplicity $d\nch/d\eta$.
  \item Identified particle yields $dN/dy$ of pions, kaons, and protons.
  \item Transverse energy production $dE_T/d\eta$.
  \item Charged particle mean transverse momentum $\langle p_T \rangle$ ($0.15 < p_T < 10$~GeV).
  \item Identified particle mean transverse momentum $\langle p_T \rangle$ of pions, kaons, and protons.
  \item Mean transverse momentum fluctuations $\delta p_T / \langle p_T \rangle$ (charged particles, $0.15 < p_T < 2.0$~GeV).
  \item Two-particle flow cumulants $\vnk{n}{2}$ for $n=2,3,4$\\ (charged particles, $0.2 < p_T < 5.0$~GeV for ALICE, and $0.3 < p_T < 3.0$~GeV for CMS).
  \item Four-particle flow cumulant $\vnk{2}{4}$ \\(charged particles, $0.2 < p_T < 5.0$~GeV).
  \item Symmetric cumulants $\SC(4, 2)$ and $\SC(3,2)$.
\end{itemize}
Each observable is calculated from the list of final state particles produced by UrQMD using the same methods applied by experiment.
We generally match the kinematic cuts of all measurements with two exceptions: we use a larger rapidity interval $|\eta| < 0.8$ than experiment for some boost-invariant observables to improve our finite particle statistics, and we do not apply a rapidity gap, e.g.\ $|\Delta \eta| > 1$, between pairs of particles when calculating the two-particle cumulant $\vnk{n}{2}$ since we already oversample particles from each hydrodynamic event, and this oversampling suppresses non-flow correlations.

@@

However, for some observables such as $p$-Pb mean $p_T$ \cite{Abelev:2013bla} and flow cumulants $\vnk{n}{k}$ \cite{Chatrchyan:2013nka}, the experiments use a special high-multiplicity trigger to select rare, ultra-central events according to the number of charged particles produced or detector tracks offline.
These high-multiplicity bins are too selective for our modest minimum-bias event sample, and so a different procedure is required.
We exploit, for this purpose, the approximate correspondence
\begin{equation}
  \label{eq:bin_equiv}
  \ntrk \sim \nch \propto N_\text{parton}
\end{equation}
between each event's initial secondary-parton density ${N_\text{parton} = \int d^2x\,n(\x)}$ in Eq.~\eqref{energy-momentum} and its final-state event activity, characterized by $\nch$ or $\ntrk$ \cite{Abelev:2013bla, Chatrchyan:2013nka}.

@@

In principle, one could calculate the likelihood function in Eq.~\eqref{eq:likelihood} directly, e.g.\ by running the model a large number of times at a given parameter point $\x$ to calculate the model observables $\y_m(\x)$ from the ensemble of simulated events, but in practice such a procedure would be intractable.
The model is computationally intensive to evaluate, and thousands of events are required to calculate the simplest observables at a single parameter point.
Moreover, we need to evaluate Eq.~\eqref{eq:likelihood} \emph{numerous} times in order to sample the multidimensional posterior distribution so that the samples may be histogrammed and visualized.

>>>>

\newcommand{\trento}{T\raisebox{-0.5ex}{R}ENTo}
\newcommand{\sqrts}{\sqrt{s_\mathrm{NN}}}
\newcommand{\fmc}{\ensuremath{\text{fm}/c}}
\newcommand{\nch}{N_\text{ch}}
\newcommand{\ntrk}{N_\text{trk}^\text{offline}}
\newcommand{\vnk}[2]{v_#1\{#2\}}
\newcommand{\sigmaf}{\sigma_\text{fluct}}
\newcommand{\X}{\chi_\text{struct}}
\newcommand{\taufs}{\tau_\text{fs}}
\newcommand{\dmin}{d_\text{min}}
\newcommand{\Tsw}{T_\text{switch}}
\newcommand{\T}{\tilde{T}}
\newcommand{\bv}{\mathbf b}
\newcommand{\xv}{\mathbf x}
\newcommand{\yv}{\mathbf y}
\newcommand{\zv}{\mathbf z}
\newcommand{\paddedhline}{\noalign{\smallskip}\hline\noalign{\smallskip}}
\newcommand{\order}[1]{$\mathcal O(10^{#1})$}
\newcommand{\note}{\textcolor{theblue}{[?]}}

@@

Consider the collision of two protons, labeled $A$ and $B$, with three-dimensional densities $\rho_A$ and $\rho_B$ in their local rest frames.
The proton-proton overlap function
\begin{equation}
  \label{eq:tpp}
  T_{pp}(b) = \int dx\, dy\, dz\, \rho_A(x, y, z)\, \rho_B(x - b, y, z),
\end{equation}
describes the eikonal overlap of the two proton wave packets at fixed impact parameter $b$, where coordinate $(x, y)$ lies in the transverse plane and $z$ is parallel to the beam axis.
Here we assume that each proton is comprised of smaller constituents---e.g.\ valence quarks, sea quarks, and small-x gluons---which may collide to produce secondary particles and contribute to the observed inelastic proton-proton cross section.

Within a picture of independent pairwise collisions between the constituents, a Glauber model model may be used to calculate the probability $P_\mathrm{coll}$ that the two protons collide inelastically at impact parameter $b$. In the limit when the number of constituents is large, it yields the particularly simple form
\begin{equation}
  \label{eq:pcoll}
  P_\mathrm{coll}(b) = 1 - \exp[-\sigma_\mathrm{eff}\, T_{pp}(b)], \\
\end{equation}
where $\sigma_\mathrm{eff}$ is an effective cross section for pairwise inelastic collisions between the constituents, and $\sigma_{pp}^\mathrm{inel}$ is the total inelastic proton-proton cross section
\begin{equation}
  \label{eq:sigma_nn}
  \sigma_{pp}^\mathrm{inel} = \int 2 \pi b\, db\, P_\mathrm{coll}(b).
\end{equation}

The proton densities $\rho_{A,B}$ in Eq.~\eqref{eq:tpp} are commonly modeled using a spherically symmetric distribution.
For instance, the original implementation of the \trento\ model uses Gaussian protons, largely because it yields a simple analytic solution to Eq.~\eqref{eq:pcoll}.
Needless to say, such approximations are admittedly crude and may have a significant effect on the dynamics of small collision systems where the proton size is comparable to the size of the produced QGP medium.
A number of previous studies have investigated the effects of deformed or ``lumpy'' protons.
One common implementation is a superposition of three valence quarks, typically described by Gaussian or exponential form factors \cite{Welsh:2016siu, Bozek:2017jog, Schenke:2014zha, Schlichting:2014ipa, Adare:2015bua, Broniowski:2016pvx}.
The corresponding proton density is then assumed to be that of predominantly small-x gluons, seeded by the distribution of color charge in each of the three valence quarks.

@@

In this work, we pursue a less restrictive and more parametric description of the proton where the number of substructure degrees of freedom is uncertain as depicted in Fig.~\ref{fig:substructure}.
We model each nucleon's three-dimensional density $\rho_{A,B}$ as a sum of $n_c$ independent constituents,
\begin{equation}
  \label{eq:rho}
  \rho_{A, B}(\xv) = \frac{1}{n_c} \sum\limits_{i=1}^{n_c} \rho_c(\xv - \xv_i),
\end{equation}
where each constituent density $\rho_c$ is described by a Gaussian distribution,
\begin{equation}
  \rho_c(\xv) = \frac{1}{(2 \pi v^2)^{3/2}} \exp\left(- \frac{|\xv|^2}{2 v^2}\right).
\end{equation}
of width $v$.
The constituent positions $\xv_i$ are sampled independently (ignoring correlations) from a Gaussian radial distribution,
\begin{equation}
  \label{eq:radial_dist}
  P(\xv_i; \xv_{A,B}) = \frac{1}{(2\pi r^2)^{3/2}} \exp\left(-\frac{|\xv_i - \xv_{A,B}|^2}{2 r^2}\right),
\end{equation}
where $\xv_{A,B}$ is the position of each proton, and $r$ is a parameter which varies the sampling radius of the constituent positions about each proton position.
As a matter of convenience, we sample the proton positions \emph{before} determining the constituent positions.
This leads to small discrepancies between the designated proton positions $\xv_{A,B}$, and the actual position of each proton's center-of-mass, owing to fluctuations in the constituent positions $\xv_i$.
The sampling radius $r$ should thus be interpreted with care; it is \emph{not} the root-mean-square radius of the sampled nucleons.

@@

The two protons $A$, $B$ are assigned a random impact parameter, and Eq.~\eqref{eq:pcoll} is used to sample their inelastic collision probability $P_\mathrm{coll}(b)$.
Note, this proton-proton inelastic collision probability has no direct knowledge of the individual constituent positions; it is only indirectly sensitive to their positions through the density $\rho_{A, B}$.
The protons in this work are described by a \emph{single} density field.
The constituents are simply used as a mechanism to generate deformity and fluctuations.
This is an important distinction between the present model and a similar nucleon substructure implementation known as the participant or ``wounded'' quark model which allows for a subset of quarks (constituents) to participate inside a single nucleon \cite{ANISOVICH1978477, Broniowski:2016pvx}.
The proton, unlike the nucleus, cannot produce semi-stable spectator fragments in a high-energy collision.
Any spectator quarks produced by a wounded quark model would be colored objects that necessarily contribute to secondary particle production as they fragment and recombine to form color-neutral hadrons.
We correspondingly require that the nucleons in Eq.~\eqref{eq:rho} participate as singular objects, such that all spectator matter discarded by the simulation is appropriately color-neutral and inert.

Hereafter, we switch to Milne coordinates $(\xv_\perp, \eta_s, \tau)$, where $\xv_\perp$ is a Cartesian coordinate $(x, y)$ in the transverse plane, $\eta_s = \tfrac{1}{2} \log[(t+z)/(t-z)]$ is the system's spacetime rapidity, and $\tau = \sqrt{t^2 - z^2}$ its proper time.
Assuming our protons collide at the sampled impact parameter $\bv$, we assign each proton a participant thickness,
\begin{equation}
  \label{eq:fluctuated_thick}
  \T_{A, B}(\xv_\perp) \equiv \int dz\, \tilde{\rho}_{A,B}(\xv_\perp \pm \bv/2, z),
\end{equation}
which projects its \emph{fluctuated} proton density,
\begin{equation}
  \tilde{\rho}_{A,B}(\xv) \equiv \frac{1}{n_c} \sum\limits_{i=1}^{n_c} \gamma_i\, \rho_c(\xv - \xv_i),
\end{equation}
onto the transverse plane $\xv_\perp$ centered at $z=0$.
This fluctuated proton density $\tilde{\rho}_{A,B}$ equals the proton density $\rho_{A,B}(\xv)$ in Eq.~\eqref{eq:rho} with additional weights $\gamma_i$ sampled from a gamma distribution with unit mean and variance $1/k$.
These \emph{ad hoc} gamma random weights are necessary to describe the large fluctuations observed in minimum bias proton-proton collisions, although their exact physical origin is not well understood.

The initial energy density $e$ deposited at midrapidity $\eta_s=0$ and proper time $\tau_0 \approx 0~\fmc$ is then some function,
\begin{equation}
  e(\xv_\perp, \eta_s=0, \tau_0) = f(\T_A, \T_B),
\end{equation}
of the local participant thicknesses $\T_A$, $\T_B$ in each nucleus.
A natural first guess for this mapping is an arithmetic mean,
\begin{equation}
  e(\xv_\perp, \eta_s=0, \tau_0) = \text{const} \times \frac{\T_A + \T_B}{2},
\end{equation}
equal to a participant or ``wounded nucleon'' model for initial energy deposition up to meaningless factor of two in the overall normalization constant.
This wounded nucleon scaling was in fact one of the first such mappings used as a proxy for initial particle production in relativistic heavy-ion collisions \cite{Bialas:1976ed}.
It was quickly realized, however, that the wounded nucleon model predicts the wrong scaling for charged particle production as a function of collision centrality and hence the wrong scaling for initial energy deposition at midrapidity as a function of participant thicknesses $\T_A$ and $\T_B$ \cite{Kharzeev:2000ph}.

A simple remedy is to replace the arithmetic mean of the wounded nucleon model with a more flexible parametrization
\begin{equation}
  \label{eq:gmean}
  e(\xv_\perp, \eta_s=0, \tau_0) = \text{const} \times \bigg( \frac{\T_A^p + \T_B^p}{2} \bigg)^{1/p},
\end{equation}
based on a family of functions known as the generalized mean(s)\footnote{Previous applications of the \trento\ model used the generalized mean ansatz to parametrize the initial entropy density $s(\xv_\perp, \eta_s=0, \tau_0)$, noting its approximate linear relation with the final charged particle density $d\nch/d\eta$ \cite{Moreland:2014oya, Bernhard:2016tnd}. In the present framework, the initial energy density $e(\xv_\perp, \eta_s=0, \tau_0)$ is approximately proportional to $d\nch/d\eta$ as well; hence, the generalized mean works equally well as a parametrization for either quantity. Here we use the energy density to make direct contact with the free-streaming equation Eq.~\eqref{energy-momentum}.}.
This parametrization introduces a dimensionless parameter $p$ which varies the scaling behavior of initial energy deposition at midrapidity.
For certain discrete values of $p$, it reduces to well known functional forms such as the arithmetic, geometric, and harmonic means:
\begin{equation}
  \newlength{\extraspace}
  \setlength{\extraspace}{0.5ex}
  e \propto
  \begin{cases}
    \max(\T_A, \T_B) & p \rightarrow +\infty, \\[\extraspace]
    (\T_A + \T_B)/2 & p = +1, \hfill \text{ (arithmetic)} \\[\extraspace]
    \sqrt{\T_A \T_B} & p = 0, \hfill \text{ (geometric)} \\[\extraspace]
    2\, \T_A \T_B/(\T_A + \T_B) & p = -1, \hfill \text{ (harmonic)} \\[\extraspace]
    \min(\T_A, \T_B) & p \rightarrow -\infty.
  \end{cases}
  \label{eq:trento_p}
\end{equation}

Conveniently, it has been shown that the generalized mean ansatz is able to mimic the scaling behavior of certain saturation based initial condition models \cite{Bernhard:2016tnd}, and hence it should serve as a reasonable parametric form for exploring QGP entropy deposition, assuming imperfect knowledge of saturation effects in nature.
Of course the model is \emph{not} a substitute for first principle theory calculations, and it may fail to reproduce nuanced features of \emph{ab initio} models such as the existence of short-range gluon field fluctuations \cite{Schenke:2012wb}.

@@

When applying the model to heavy-ions, we sample nucleon positions from a Woods-Saxon density distribution subject to a minimum distance criteria $|\xv_i - \xv_j| > d_\mathrm{min}$ between all pairs of nucleons $i$, $j$.
The minimum distance algorithm, first described in Ref.~\cite{Bernhard:2018hnz}, uses a simple trick to resample the nucleon positions without modifying the target Woods-Saxon radial distribution.
We first presample the radii of all nucleons in a given nucleus and sort them in ascending order.
We then sample the solid angles of each nucleon one-by-one, starting with the nucleon closest to the center of the nucleus and working our way outwards.
If a sampled nucleon position is too close to any of its previously placed neighbors, its solid angle is resampled until the minimum distance criteria is satisfied.
Similar methods could be used to model correlations between individual constituents inside each nucleon, although the numerical implementation would be somewhat tedious.

@@

The nuclear collision model constructed in Sec.~\ref{sec:model} includes a number of free parameters $\xv$ which describe the initial state, pre-equilibrium dynamics, and hydrodynamic medium.
Given values for the parameters $\xv$, the model may be used to predict a vector of simulated observables $\yv_m$.
For example, $\yv_m$ might be a vector consisting of charged particle yields in different centrality bins.
The physics model thus represents a vector-valued function $f(\xv) = \yv_m$ which maps the parameter values $\xv$ to the simulated observables $\yv_m$.

The goal of this work is to estimate the true model parameters $\xv_\star$ provided some evidence that our model predictions $\yv_m$ describe experimental measurements $\yv_e$.
The problem involves three distinct components:
\begin{enumerate}[itemsep=0pt, leftmargin=2\parindent]
  \item $H_f$: the hypothesis that the nuclear collision model $f$ formulated in this work provides a realistic description of reality,
  \item $H_\xv$: the hypothesis that the model parameters $\xv$ are the true model parameters $\xv_\star$ of $f$, and
  \item $E$: the evidence provided by the experimental data $\yv_e$ and its corresponding uncertainties.
\end{enumerate}
As a practical matter, we always assume that hypothesis $H_f$ is correct, meaning there are no glaring flaws in our chosen theoretical framework.
This is a significant assumption; the application of hydrodynamic simulations to small collision systems is speculative, and our conclusions are conditional on the framework making sense.

Subject to this assumption, we can apply Bayes' theorem to evaluate the hypothesis $H_\xv$ for the true model parameters,
\begin{equation}
  \label{eq:bayes}
  P(H_\xv | E) \propto P(E | H_\xv)\, P(H_\xv).
\end{equation}
The left-side of this expression is the \emph{posterior}: the probability for hypothesis $H_\xv$ given the experimental evidence $E$.
On the right-side there are two separate terms.
The first term $P(E | H_\xv)$ is the \emph{likelihood} function: the probability of observing the experimental evidence $E$ given our model and the hypothesis $H_\xv$ for the true model parameters $\xv_\star$, and the second term $P(H_\xv)$ is the \emph{prior}: an estimate of the probability of hypothesis $H_\xv$ in the absence of evidence $E$.

We assume that the likelihood function in Eq.~\eqref{eq:bayes} is described by a multivariate normal distribution
\begin{equation}
  \label{eq:likelihood}
  P(E | H_\xv) = \frac{1}{\sqrt{(2\pi)^m \det \Sigma}} \exp \left ( -\frac{1}{2}\Delta\yv^\intercal \Sigma^{-1} \Delta\yv \right ),
\end{equation}
where $\Delta\yv = \yv_m(\xv) - \yv_e$ is the discrepancy of the model and experiment, and $\Sigma = \Sigma_m(\xv) + \Sigma_e$ is the \emph{total} covariance matrix, equal to the sum of a modeling component $\Sigma_m(\xv)$ and an experimental component $\Sigma_e$ which account for all known sources of uncertainty in the simulated and measured observables.

@@

For the prior $P(H_\xv)$, we specify ranges, i.e.\ minimum and maximum values, for each parameter which are listed in Table~\ref{tab:design}.
We assume the prior distribution is constant and nonzero within each specified range and zero otherwise.
The selected parameter ranges are intentionally wide to avoid clipping the calibrated posterior; for example, a previous analysis of the \trento\ model \cite{Bernhard:2016tnd} found $p \sim 0$, but we use a prior range $p \in [-1, 1]$ to account for differences in the present model, e.g.\ nucleon substructure, which could modify its posterior.
Several of the model parameters require special care and are reparametrized accordingly:

@@

The likelihood function \eqref{eq:likelihood} provides evidence for \mbox{(or against)} the model parameters $\xv$ by comparing the model predictions $\yv_m(\xv)$ to experimental data $\yv_e$.
We focus on simple experimental observables in the present study which are sensitive to the bulk properties of the produced medium.
We calculate for each set of model parameters the following observables at midrapidity:
\begin{itemize}[leftmargin=1\parindent, itemsep=0pt]
  \item Charged-particle multiplicity $d\nch/d\eta$.
  \item Identified particle yields $dN/dy$ of pions, kaons, and protons.
  \item Transverse energy production $dE_T/d\eta$.
  \item Charged particle mean transverse momentum $\langle p_T \rangle$ ($0.15 < p_T < 10$~GeV).
  \item Identified particle mean transverse momentum $\langle p_T \rangle$ of pions, kaons, and protons.
  \item Mean transverse momentum fluctuations $\delta p_T / \langle p_T \rangle$ (charged particles, $0.15 < p_T < 2.0$~GeV).
  \item Two-particle flow cumulants $\vnk{n}{2}$ for $n=2,3,4$\\ (charged particles, $0.2 < p_T < 5.0$~GeV for ALICE, and $0.3 < p_T < 3.0$~GeV for CMS).
  \item Four-particle flow cumulant $\vnk{2}{4}$ \\(charged particles, $0.2 < p_T < 5.0$~GeV).
  \item Symmetric cumulants $\SC(4, 2)$ and $\SC(3,2)$.
\end{itemize}
Each observable is calculated from the list of final state particles produced by UrQMD using the same methods applied by experiment.
We generally match the kinematic cuts of all measurements with two exceptions: we use a larger rapidity interval $|\eta| < 0.8$ than experiment for some boost-invariant observables to improve our finite particle statistics, and we do not apply a rapidity gap, e.g.\ $|\Delta \eta| > 1$, between pairs of particles when calculating the two-particle cumulant $\vnk{n}{2}$ since we already oversample particles from each hydrodynamic event, and this oversampling suppresses non-flow correlations.

@@

However, for some observables such as $p$-Pb mean $p_T$ \cite{Abelev:2013bla} and flow cumulants $\vnk{n}{k}$ \cite{Chatrchyan:2013nka}, the experiments use a special high-multiplicity trigger to select rare, ultra-central events according to the number of charged particles produced or detector tracks offline.
These high-multiplicity bins are too selective for our modest minimum-bias event sample, and so a different procedure is required.
We exploit, for this purpose, the approximate correspondence
\begin{equation}
  \label{eq:bin_equiv}
  \ntrk \sim \nch \propto N_\text{parton}
\end{equation}
between each event's initial secondary-parton density ${N_\text{parton} = \int d^2x\,n(\xv)}$ in Eq.~\eqref{energy-momentum} and its final-state event activity, characterized by $\nch$ or $\ntrk$ \cite{Abelev:2013bla, Chatrchyan:2013nka}.

@@

In principle, one could calculate the likelihood function in Eq.~\eqref{eq:likelihood} directly, e.g.\ by running the model a large number of times at a given parameter point $\xv$ to calculate the model observables $\yv_m(\xv)$ from the ensemble of simulated events, but in practice such a procedure would be intractable.
The model is computationally intensive to evaluate, and thousands of events are required to calculate the simplest observables at a single parameter point.
Moreover, we need to evaluate Eq.~\eqref{eq:likelihood} \emph{numerous} times in order to sample the multidimensional posterior distribution so that the samples may be histogrammed and visualized.
